Live Functional Programming with Typed Holes

### 2018-02-21 examples.tex
Section 2, first full rough draft

\begin{lstlisting}
sumList : list(int) -> int
sumList [] = ?[]?
sumList (n:ns) = n + sumList ns
\end{lstlisting}

\noindent
%
Because the list expression on line 2 does not have type \li{int} as required,
it is wrapped in a (non-empty) hole by the bidirectional type
checker~\cite{popl-paper}.
%
Rather than trying to debug the error based on the static error, the programmer
may wish to trying running the function anyway by calling, say, \li{sumList(2)}.
%
\HazelnutLive{} runs and produces the indeterminate expression \li{3 + ?[]?}.
%
By observing that the hole expression is being added to the integer \li{3}, he
realizes that it needs to be an integer, specifically, \li{0}.
%
Compared to the trace displayed by \citet{Seidel2016}, the indeterminate result
produced by \HazelnutLive{} is ``flattened'' because the expression \li{1 + 2}
successfully proceeded to evaluate despite the error elsewhere.

@@

An alternative to marking type-inconsistent expressions as sources of error
(with non-empty holes) is to use \emph{gradual types}~(\eg{}~\cite{XXX,XXX,XXX})
to sidestep the choice of a assigning single static type, instead deferring to
dynamic casts.
%
\HazelnutLive{} inherits the notion of \emph{type holes} from
\citet{popl-paper}, which serves a similar static purpose as the unknown type in
gradual type systems.

@@

Consider the \li{negate} function (adapted from \cite{ChughPOPL2012}), which
does not have a conventional static type (without using expression holes),
because the argument \li{x} is used at type \li{int} on line \rkc{XXX} and
function type on line \rkc{XXX}.
%
Therefore, the declared type of \li{x} is the hole type \li{??}, which allows
\li{x} to be used at conflicting types by expanding each use with \emph{casts}
that ensure the safety of each dynamic use.
%
The return type also uses the hole type \li{??}.

@@

\noindent
%
As in prior gradually typed languages~\cite{XXX},
evaluating the expression \li{negate true 1} on line \rkc{XXX} leads to
\li{(not (1<int => ??><?? => bool>))<bool => ??>}, where the inner casts
lead to the \emph{cast error} \li{1<?? =/=> bool>} because \li{1} cannot be
safely treated as having type \li{bool}.
%
Unlike in prior gradually typed languages, however, \HazelnutLive{} evaluates
around the cast error, making progress on the \li{2 + 3 + negate false 4}
expression that surrounds the error; the final indeterminate result is
\li{(not (1<?? =/=> bool>))<bool => ??> + 1}.
%
Just like it is useful for static type checkers to report multiple errors, our
approach allows us to report multiple dynamic cast errors, and otherwise make
progress on expressions that do not depend on failed casts.
%
This approach can be incorporated into existing gradually typed languages.

\rkc{STOPPING HERE FOR NOW}

@@

\overviewExample{6}{\rkc{...}}

\rkc{breakpoints, understanding input/output behavior}

\parasection{Example 6: Quicksort}

\begin{lstlisting}
quicksort : list('a) -> list('a)
quicksort [] = []
quicksort (x:xs) =
  let (left, right) = span ((>) x) xs in
  ??
\end{lstlisting}

programmer has some intended relationship between names and values they bind.
%
live results show that left has larger elements and right has smaller.
%
right away, can see that need to change filter predicate. (alternatively, the
names could have been swapped, but that order of names was the natural way
to think about it.)
%
but next, still see that right has some smaller elements.
%
should pick partition instead of span.

\begin{lstlisting}
quicksort : list('a) -> list('a)
quicksort [] = []
quicksort (x:xs) =
  let (left, right) = partition ((<) x) xs in
  let (left', right') = (quicksort left, quicksort right) in
  ? (left' @ [x] @ right) ?
\end{lstlisting}

now, put the lists back together.
%
doesn't do what is expected.
%
live feedback shows right' is computed correctly, but ah, it's not being used
in the return list.
%
fix, and re-run, keeping the final expression in a hole.
%
can visualize that the nested appends are being built up correctly, and that the
result would be the correct list.
%
finalize the hole and resume. everything but the nested appends can be reused.
%
\emph{hole environment traces} or \emph{hole environment traces}

>>>>

\begin{lstlisting}
sum_list : list(int) -> int
sum_list [] = ?[]?
sum_list (n:ns) = n + sum_list ns
\end{lstlisting}

\noindent
%
Because the list expression on line 2 does not have type \li{int} as required,
it is wrapped in a (non-empty) hole by the bidirectional type
checker~\cite{popl-paper}.
%
Rather than trying to debug the error based on the static error, the programmer
may wish to trying running the function anyway by calling, say, \li{sum_list(2)}.
%
\HazelnutLive{} runs and produces the indeterminate expression \li{3 + ?[]?}.
%
By observing that the hole expression is being added to the integer \li{3}, he
realizes that it needs to be an integer, specifically, \li{0}.
%
Compared to the trace displayed by \citet{Seidel2016}, the indeterminate result
produced by \HazelnutLive{} is ``flattened'' because the expression \li{1 + 2}
successfully proceeded to evaluate despite the error elsewhere.

@@

\emph{Gradual type systems}~(\eg{}~\cite{XXX,XXX,XXX}) can statically accept
programs that would otherwise be rejected by static type systems---either
because type inference cannot reconstruct a valid type assignment, or because
there may not be a unique valid type assignment at all.
%
In either case, gradual type systems allow types to contain \emph{unknown
types}, and partially unknown types can be used in contexts that require
different types, so long as they are \emph{consistent} (intuitively, equal
modulo the unknown parts).
%
Dynamic casts are then inserted to ensure that these remaining static
discrepancies are not violated at run-time.

\HazelnutLive{} inherits the notion of \emph{type holes} from
\citet{popl-paper}, which serves a similar static purpose as the unknown type in
gradual type systems.
%
In contrast to prior gradually typed languages, however, \HazelnutLive{} can
evaluate ``around'' cast errors in the same way as it does for expression holes.

@@

Consider the \li{negate} function (adapted from \cite{ChughPOPL2012}), which
cannot be assigned a static type in a conventional type system---either a
bidirectional one, like in \HazelnutLive{}, or one with ML-style,
unification-based inference---because the argument \li{x} is used at type
\li{int} on line \rkc{XXX} and \li{bool} on line \rkc{XXX}.
%
Therefore, the declared type of \li{x} is the hole type \li{??}, which allows
\li{x} to be used at the conflicting types, with each use expanded into an
expression wrapped in a \emph{cast} that will dynamically check for safety.
%
The return type also uses the hole type \li{??}, leading to additional casts in
the expansion.

@@

\noindent
%
As in prior gradually typed languages~\cite{XXX},
evaluating the expression \li{negate false 1} on line \rkc{XXX} leads to
\li{(not (1<int => ??><?? => bool>))<bool => ??>}, where the inner casts
lead to the \emph{cast error} \li{1<?? =/=> bool>} because \li{1} cannot be
safely treated as having type \li{bool}.
%
Unlike in prior gradually typed languages, however, \HazelnutLive{} evaluates
around the cast error, making progress on the \li{2 + 3 + negate false 4}
expression that surrounds the error; the final indeterminate result is
\li{(not (1<?? =/=> bool>))<bool => ??> + 1}.
%
Just like it is useful for static type checkers to report multiple errors, our
approach allows us to report multiple dynamic cast errors, and otherwise make
progress on expressions that do not depend on failed casts.
%
This approach can be incorporated into existing gradually typed languages.

@@

Previous examples highlighted how running different kinds of incomplete
programs---ones with missing expressions, type-inconsistent expressions, or
missing types---can be useful.
%
Our final example demonstrates how expression holes may be useful even for
complete programs, in order to facilitate debugging and program understanding.

%% \overviewExample{6}{\rkc{...}}
%%
%% \rkc{breakpoints, understanding input/output behavior}

\overviewExample{6}{Quicksort}

Consider the following buggy implementation of quicksort, which contains several
errors.
%
The student wants to debug the implementation and so inserts an expression
around the return expression on line \rkc{XXX}, and runs
\li{quicksort [X,X,X,X,X,X]}.

\begin{lstlisting}
quicksort : list('a) -> list('a)
quicksort [] = []
quicksort (x:xs) =
  let (left, right) = span ((>) x) xs in
  let (left', right') = (quicksort left, quicksort right) in
  ? (left' @ [x] @ right) ?
\end{lstlisting}

\noindent
%
Based on the live environment for the initial call, the student notices that the
values in \li{left} are larger than the pivot \li{x}, but \li{left} was intended
to bind smaller values.
%
This suggests that the student got the filtering predicate backwards, so he
replaces it with \li{(<)} on line \rkc{XXX}.
%
After re-running the program and viewing the live output again, \li{left} now
contains smaller values than \li{x}, but not all of them; \li{right} has some
values that are larger and some that are smaller.
%
The student uses type-based search to find other functions, besides \li{span},
of type \li{('a -> bool) -> 'a list -> ('a list, 'a list)} and finds
\li{partition}.
%
It may be easier for the student to simply try the function rather than studying
the documentation, so he edits line \rkc{XXX} to call \li{partition} instead,
re-runs, and observes that both \li{left} and \li{right} now seem to achieve the
intended partitioning.

Despite these two fixes, the result expression is not yet sorted.
%
In particular, the values greater than the pivot \li{x} are not completely
sorted.
%
Viewing the values of bindings of \li{left'} and \li{right'}, the student sees
that both seem to have correct prefixes of sorted lists, but that they are
incorrect after the pivots.
%
This suggests that something may be wrong with the handling of \li{right'}.
%
Indeed, the student takes a look for uses of \li{right'} and sees that line
\rkc{XXX} mistakenly uses \li{right} instead.
%
Making this change and re-running now produces the desired result, built up as a
hole environment tree that shows how all of the recursive results will be
combined by a series of nested appends.
%
Happy with the results of the program, the student removes the hole around the
expression on line \rkc{XXX}, and evaluation can proceed to perform the nested
appends, producing the final sorted result.
%
\autoref{sec:discussion} discusses how debugging tools may be designed to
systematically insert hole expressions in particular places to achieve certain
debugging and program understanding strategies.
%
%% When run on a larger input, this projection of the full evaluation trace shows a
%% nesting structure---which could be rendered as visualization to help
%% understanding, as described in \autoref{sec:discussion}---whose depth is clearly
%% much smaller than the length of the input list; this suggests that the
%% algorithm, indeed, achieves some sort of sublinear complexity, as intended.
%

\begin{lstlisting}
quicksort : list('a) -> list('a)
quicksort [] = []
quicksort (x:xs) =
  let (left, right) = partition ((<) x) xs in
  let (left', right') = (quicksort left, quicksort right) in
  ? (left' @ [x] @ right') ?
\end{lstlisting}
